Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-01-22 paper/paper.tex
added brzozowski axioms

\section{Type-theoretic Characterizations of Language Classes}
Classically, formal language theory is closely related to automata theory. In the styles of Chomsky and Sch\"utzenberger, there is a hierarchy of language classes and the types of automata that recognize them. The type theory of \cref{sec:typetheory} serves as a syntax to reason about formal grammars; moreover, we describe notions of automata into this formal system via syntactic forms that describe an automaton as a grammar. This method allows us to internalize many classic automata-theoretic results as proof terms in our type theory.

@@

Let $\hat{Y}_{k}$ refer to the syntactic term on the right hand side of the definition of $Y_{k}$ inside of an NFA $N$. We say that $N$ is a \emph{deterministic finite automaton} (DFA) if there are no $\epsilon$-transitions and for every definition $\hat{Y}_{k}$ the $c_{j}$ are distinct guards for each produced state.
% \steven{Is this the right notion of determinism? Perhaps, it is is better to define determinism as something like: there is at most one derivation for all $j, k$, for all $a \in \Sigma$ we have $x : Y_{k}, y : a \vdash z : Y_{j}$.}

@@

The class of \emph{regular} grammars is given as follows:

\[
Reg ::= 0 ~|~ I ~|~ c ~|~ Reg \oplus Reg ~|~ Reg \otimes Reg ~|~ Reg^{*}
\]
where $c \in \mathcal{A}$ is a character literal drawn from a fixed alphabet, and the Kleene star $\cdot^{*}$ is derived as an admissible operation using the fixed point operator --- $g^{*} = \mu X.I \oplus (g \otimes X)$. \steven{This definition is maybe fair to assume. Also maybe better suited for a section earlier in the paper}

\begin{theorem}
  \label{thm:thompson}
  For $g$ a regular grammar, there is an NFA $N$ that recognizes the same language as $g$. Further, $N$ and $g$ are isomorphic.
\end{theorem}

@@

where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively. This finshes the construction of a parse transformer $NFA(g \oplus g') \to g \oplus g'$ The map in the other direction is given the derivation,

@@

For this taylor iso to hold, we need the map $c \otimes (c \lto A) \vdash (c \otimes \top) \amp A$ to be an iso. Specifically, this term needs an iso $\phi : c \otimes (c \lto A) \to (c \otimes \top) \amp A$ defined below,

@@

From here we need to also show that derivatives of regular languages are regular (look at Brzozowksi's proof) and that regular languages have finitely many derivatives (potentially with some size constraint on the derivative?). Then the last iso in the chain above can be used to build a DFA equivalent to $A$ using Brzozowksi's algorithm.

@@

\section{NFA stuff}
\label{sec:appendixnfa}

>>>>

\section{Type Theoretic Formal Grammar Theory}

Now that we have developed a type theoretic syntax for formal grammars
and parse transformers, we show how many common concepts of formal
language theory and grammars are naturally expressed using type
theoretic notions, and some constructions and proofs can be carried
out in our type theory as well.

Firstly, regular expressions and their extension to context-free
expressions are naturally expressed as a limited syntax of linear
types. That is, regular expressions are precisely the linear types
constructed using only $c, I, \otimes, 0, \oplus$ and the Kleene star,
which can be defined as $A^* = \mu x. I \oplus A \otimes X$. The
context-free expressions are similarly expressible by allowing not
just the Kleene star, but unrestricted usage of $\mu$. Remember,
however, that these are a notion of regular expressions that maintain
a notion of ambiguous grammar. That is, $A \oplus A$ is not considered
to be the same as $A$. While there does not appear to be an analogous
weakening of this definition to naturally describe the
context-sensitive languages, we will see shortly that the indexing of
linear types by non-linear types allows for capturing a notion of
context in grammars.

In addition to expressing algebraic presentations of grammars, the
type-theoretic syntax allows for naturally expressing grammars that
capture the behavior of automata. \max{sync this up with Steven's
  stuff, and add stuff about PDAs and Turing Machines}.

A key construct in formal language theory is the Brzozowski derivative
\cite{brzozowski}. Given a string $w$ the (left) derivative of a
language $L$ with respect to $w$, written $\partial_w L$ is the language
\[ (\partial_wL ) = \{ ww' \pipe w' \in L \}\]
This semantics is given using the linear implication $w \tol
L$. Expressing Brzozowski's calculation of the derivative of a regular
expression relies on additional axioms to the type theory.\max{axioms
  to carry out Brzozowski's derivative calculations plz.}

\steven{The axioms that I think we need:

$c, c'$ characters. $g$ any grammar.

\begin{gather}
  c \lto c \cong I \\
  c \lto c' \cong 0, c \neq c' \\
  c \lto I \cong 0 \\
  c \lto g \otimes h \cong \left( (c \lto g) \otimes h \right) \oplus \left( (g \amp I) \otimes (c \lto h) \right) \\
  I \lto c \cong c \\
  c \lto (g \oplus h) \cong (c \lto g) \oplus (c \lto h) \\
  c \lto (c' \lto g) \cong (c' \otimes c) \lto g
\end{gather}

For point $(5)$, Brzozowski has $I \lto g \cong g$ for all grammars $g$, but I don't think that holds for us because we want to preserve ambiguity. We could instead have something like $I \lto g \cong (g \amp I)^{*} \otimes g$. For our purposes the character presentation seems sufficient, as the derivative reasoning Brzozowski uses only takes derivatives with respect to strings.

We want $c \lto g^{*} \cong (g \amp I)^{*} \otimes (c \lto g) \otimes g^{*}$, but this is derivable utilizing axiom $(4)$ and $g^{*} \cong I \oplus (g \otimes g^{*})$.

To make the DFA proof go through I think we additionally need the map $c \otimes (c \lto A) \vdash (c \otimes \top) \amp A$ to be an iso. Specifically, this term needs an iso $\phi : c \otimes (c \lto A) \to (c \otimes \top) \amp A$ defined below,

}

Using the $G$ modality, we can internalize some of the external
reasoning about grammars into the language of non-linear types. Recall
that the denotation of $G A$ is the set of parses of the empty
string. As is well known in formal language theory, all the parsing
behavior of a grammar can be reduced to reasoning about parsing of the
empty string by using the derivative of a grammar. For instance, the
set of parses for a fixed string $w$ can be defined as $G(w \tol A)$
and so the type of all parses of a grammar can be defined as
$\Sigma(w:\textrm{String}) G(w \tol A)$. This can be used to define
internally a notion of \emph{ambiguity} of a grammar. A grammar $A$ is
ambiguous \steven{This is an unambiguous grammar, right?} if for every string $w$, the type $G(w \tol A)$ of
$w-$parses is a \emph{proposition} in the sense that all terms in $G(w
\tol A)$ are equal.\max{I think we should add a propositional
  truncation to the type theory.}  Similarly, the type of \emph{total
parsers} can be defined as $G(\top \tol A)$. A total parser encodes a
function that selects a parse for every possible string. A
\emph{semi-parser} is then defined as $G(\top \tol A \oplus \top)$,
where the parser is allowed to fail for any reason. This has the
downside that the always failing semi-parser $\lambda _. \textrm{in}_r
()$ is valid for any grammar. If we were working with a full
\emph{bunched} type theory, we would be able to express the more
precise type of a \emph{deciding} parser as $G(\top \tol A \oplus \neg
A)$ where $\neg A = A \Rightarrow 0$ uses the intuitionistic
implication.

Classically, formal language theory is closely related to automata theory. In the styles of Chomsky and
Sch\"utzenberger, there is a hierarchy of language classes and the types of automata that recognize them. Inside of the
type theory of \cref{sec:typetheory}, we can characterize these language classes; moreover, we demonstrate equivalence
of each class to the associated automaton as an isomorphism within our logic.

@@

Let $\hat{Y}_{k}$ refer to the syntactic term on the right hand side of the definition of $Y_{k}$ inside of an NFA $N$.
We say that $N$ is a \emph{deterministic finite automaton} (DFA) if there are no $\epsilon$-transitions and for every
definition $\hat{Y}_{k}$ the $c_{j}$ are distinct guards for each produced state.

% \steven{Is this the right notion of determinism? Perhaps, it is is better to define determinism as something like: there is at most one derivation for all $j, k$, for all $a \in \Sigma$ we have $x : Y_{k}, y : a \vdash z : Y_{j}$.}

@@

% The class of \emph{regular} grammars is given as follows:

% \[
% Reg ::= 0 ~|~ I ~|~ c ~|~ Reg \oplus Reg ~|~ Reg \otimes Reg ~|~ Reg^{*}
% \]
% where $c \in \mathcal{A}$ is a character literal drawn from a fixed alphabet, and the Kleene star $\cdot^{*}$ is derived as an admissible operation using the fixed point operator --- $g^{*} = \mu X.I \oplus (g \otimes X)$. \steven{This definition is maybe fair to assume. Also maybe better suited for a section earlier in the paper}

\begin{theorem}
  \label{thm:thompson}
  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same language as $g$. Further, $N$ and $g$ are isomorphic.
\end{theorem}

@@

where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively.
This finshes the construction of a parse transformer $NFA(g \oplus g') \to g \oplus g'$.
The map in the other direction is given the derivation,

@@

@@

From here we need to also show that derivatives of regular languages are regular (look at Brzozowski's proof) and that regular languages have finitely many derivatives (potentially with some size constraint on the derivative?). Then the last iso in the chain above can be used to build a DFA equivalent to $A$ using Brzozowski's algorithm.

@@

\section{Equivalence Betweeen Regular Grammars and Finite Automata}
\label{sec:appendixnfa}
